Blocks of extended ancestry around selected loci in hybrid zones

### 2016-02-22 drafts/hybrid_zone.tex
.

\documentclass[11pt,letterpaper]{article}
\usepackage[margin=1in]{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{color}
\usepackage{amssymb}
\usepackage{setspace}

@@

\abstract{Hybrid zones formed between recently diverged populations offer an opportunity to better understand the mechanisms underlying reproductive isolation and the process of speciation. \yb{Some form of selection against hybrid genotypes is likely common in many hybrid zones. Here, we use a combination of analytical theory and explicit forward simulations} to gain a better understanding of how this impacts patterns of introgression \yb{across genomic and geographic} space. \yb{We find that ancestral blocks surrounding single-locus incompatibilities} are longer than average \yb{ancestry blocks}, especially when removed from the center of the hybrid zone.  The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.}

@@

\section*{Introduction}
The process of speciation commonly involves allopatric divergence in relative isolation \citep{Coyne2004}.  
When formerly isolated populations come back into contact \yb{before reproductive isolation is complete, gene flow is still possible. Thus, \alisa{ancestries} originating from each of the two diverged populations spread from their point of contact across a geographic hybrid zone --- establishing a gradient of ancestry across geography, \citep{Barton1985}}. 
%In addition to steep clines in ancestry,
\yb{The recent exchange of genetic material between diverged populations leaves a distinct pattern of variation across the genome, in which long tracts of divergent haplotypes from each ancestral population  are broken up by historical recombination events.}

@@

\paragraph{Establishment of  \alisa{a cline at the selected locus}}
Our simple model of underdominance considers a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline \citep{Barton1979}.
To study gene flow past this cline, we \alisa{turn our attention to} how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring \alisa{(i.e. the variance of the distribution of dispersal distances)}.
Assuming that: (i) alleles locally assort into diploids randomly, (ii) habitat and dispersal are homogeneous, and (iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is

\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{Bazykin1969}; for discussion of diploidy see \citep{diploidcline} \alisa{not sure which reference this is}.
Although there is not an exact analytic solution to this equation, a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. This relies on numerous approximations, but importantly, the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially. Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} does not affect this solution, and so the cline establishes over a timescale of the order $1/s$. In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. \alisa{I don't think I understand the independence from $\sigma$?}

@@

@@

\section{Nuemrical solutions for haplotype probabilities}
\label{apx:haplotype_calcs}.

>>>>

\documentclass[11pt,letterpaper]{article}
\usepackage[margin=1in]{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{color}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}

@@

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity to better understand the mechanisms underlying reproductive isolation and the process of speciation. \yb{Some form of selection against hybrid genotypes is likely common in many hybrid zones. Here, we use a combination of analytical theory and explicit forward simulations} to gain a better understanding of how this impacts patterns of introgression \yb{across genomic and geographic} space. \yb{We find that ancestral blocks surrounding single-locus incompatibilities} are longer than average \yb{ancestry blocks}, especially when removed from the center of the hybrid zone.  The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
}

@@

Main points:
\begin{itemize}
    \item 
\end{itemize}

%%%%% %%%%%% %%%%%%%
\section*{Introduction}

% summary of intro:
%  clines happen
%  but then decay unless selection
%  which affects haplotypes and LD
%  most work so far just looks at clines
%  but haplotype blocks should be helpful
%  here we look at that stuff

The process of speciation commonly involves allopatric divergence in relative isolation \citep{Coyne2004}.  
When formerly isolated populations come back into contact 
before reproductive isolation is complete, gene flow is still possible. 
As two such populations interbreed,
a gradient

Thus, \alisa{ancestries} originating from each of the two diverged populations spread from their point of contact across a geographic hybrid zone 
--- establishing a gradient of ancestry across geography, \citep{Barton1985}. 
%In addition to steep clines in ancestry,
\yb{The recent exchange of genetic material between diverged populations leaves a distinct pattern of variation across the genome, in which long tracts of divergent haplotypes from each ancestral population  are broken up by historical recombination events.}

@@

\paragraph{Establishment of  \alisa{a cline at the selected locus}}
Our simple model of underdominance considers a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline \citep{Barton1979}.
To study gene flow past this cline, we \alisa{turn our attention to} how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring \alisa{(i.e. the variance of the distribution of dispersal distances)}.
Assuming that: (i) alleles locally assort into diploids randomly, (ii) habitat and dispersal are homogeneous, and (iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{Bazykin1969}; for discussion of diploidy see \citep{diploidcline} \alisa{not sure which reference this is}.
Although there is not an exact analytic solution to this equation, a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. This relies on numerous approximations, but importantly, the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially. Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} does not affect this solution, and so the cline establishes over a timescale of the order $1/s$. In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. \alisa{I don't think I understand the independence from $\sigma$?}

@@

For completeness, here we provide one possible derivation of equation \eqref{eqn:cline_pde},
which gives the expected time evolution of $p(x,t)$
under a particular concrete model.
Other derivations are given in \citet{thesepapers};
see especially \citet{thatone}, which does not assume locally random mating.
A derivation in the same style is found in \citet{ralph2015patchy}.

For concreteness, we will discuss a population of outcrossing plants.
Suppose that the population is composed of a finite number of discrete demes,
each having $N$ diploid individuals.
Suppose that each individual at $x$ chooses to reproduce at random, 
exponentially distributed intervals,
producing a random number $X$ of seeds
using pollen coming from a random location $x+R_p$.
Each seed then disperses to a random location $x+R_s$,
where it replaces a randomly chosen extant individual.
Suppose that

Let $\Phi_{AA}(x,t)$ be the proportion of individuals at location $x$ and time $t$
with genotype $AA$, and likewise for $\Phi_{AB}$ and $\Phi_{BB}$.
\begin{align*}
    \frac{\partial}{\partial t}
    \E[ f(\Phi_t^N) ]
    &=
    N \sum_x
    \left\{
        \E\left[ 
            \Phi_t(x-R_s) \Phi_t(x-R_s-R_p)  # AA

            + \left( \Phi_t(x-R_s) (1-\Phi_t(x-R_s-R_p)) + (1-\Phi_t(x-R_s)) \Phi_t(x-R_s-R_p) \right)
            + (1-\Phi_t(x-R_s)) (1-\Phi_t(x-R_s-R_p))
        \right]
    \right\}
\end{align*}

@@

\section{Numerical solutions for haplotype probabilities}
\label{apx:haplotype_calcs}.
